---
title: "Lab 04 - Data Visualization"
output:
  html_document:
    html_preview: false
link-citations: yes
---



```{r setup, message=FALSE, warning=FALSE}
library(data.table)
library(leaflet)
library(tidyverse)
```

# Deliverables

- Answer questions below, knit to html and submit to Quercus.

# Learning Goals

- Read in and prepare the meteorological dataset
- Create several graphs with different `geoms()` in `ggplot2`
- Create a facet graph
- Conduct some customizations of the graphs
- Create a more detailed map using `leaflet()`


# Lab Description

We will again work with the meteorological data from last week.

**The objective of the lab is to examine the association between weekly average dew point temperature and wind speed in four regions of the US and by elevation.**

# Steps

### 1. Read in the data

First download and then read in with data.table:fread()

```{r, echo=TRUE, message=FALSE, eval=FALSE}
download.file(
  "https://raw.githubusercontent.com/JSC370/JSC370-2025/main/data/met/met_all.gz",
  destfile = "met_all.gz",
  method   = "curl",
  timeout  = 60
  )

met <- data.table::fread("met_all.gz")
```

### 2. Prepare the data.

- Based on last week's lab, remove suspect/incorrect observations.
- Generate a date variable.
- Using the `data.table::week` function, keep the observations of the first week of the month.
- Compute the mean by station of the variables `temp`, `rh`, `wind.sp`, `vis.dist`, `dew.point`, `lat`,`lon`, and `elev`.
- Create a categorical variable for elevation (break point at 252 m)
- Create a region variable for NW, SW, NE, SE based on lon = -98.00 and lat = 39.71 degrees

```{r}

```

### 3. Use `geom_violin` to examine the wind speed and dew point temperature by region.

You saw how to use `geom_boxplot` in class. Try using `geom_violin` instead (take a look at the help). Hint: You will need to set the `x` aesthetic to 1.

- Use facets.
- Make sure to deal with `NA` category.
- Describe what you observe in the graph.

```{r}

```


### 4. Use `geom_jitter` with `stat_smooth` to examine the association between dew point temperature and wind speed by region.

- Color points by region
- Make sure to deal with `NA` category
- Fit a linear regression line by region
- **Describe** what you observe in the graph

```{r}

  
```


### 5. Use `geom_bar` to create barplots of the weather stations by elevation category coloured by region.

- Bars by elevation category using `position = "dodge"`
- Change colors from the default. Color by region using `scale_fill_brewer` see [this](http://rstudio-pubs-static.s3.amazonaws.com/5312_98fc1aba2d5740dd849a5ab797cc2c8d.html)
- Create nice labels on axes and add a title
- Try a different theme
- Describe what you observe in the graph
- Make sure to deal with `NA`

```{r}


```

### 6. Use `stat_summary` to examine mean dew point and wind speed by region with standard deviation error bars.

- Make sure to remove `NA`
- Use fun.data="mean_sdl" in `stat_summary`
- Add another layer of `stats_summary` but change the geom to `"errorbar"` (see the help).
- Describe the graph and what you observe

```{r}

```

- Dew point temperature is...
- Wind speed is...

### 7. Make a map showing the spatial trend in relative humidity (`rh`) in the US

- Make sure to remove `NA`
- Use `leaflet()`
- Make a colour palette with custom colours
- Use `addMarkers` to include the top 10 places in relative h (hint: this will be useful `rank(-rh) <= 10`)
- Add a legend
- Describe trend in RH across the US

```{r}

```


### 8. Use a ggplot extension

- Pick and extension (except cowplot) from [here](https://exts.ggplot2.tidyverse.org/gallery/) and make a plot of your choice using the `met` data (or `met_avg`)
- Might want to try examples that come with the extension first (e.g. `ggtech`, `gganimate`, `ggforce`)

