---
title: "Lab 11 - Interactive Visualization"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval = FALSE, include  = TRUE)
```

# Learning Goals

- Read in and process Starbucks data.
- Create interactive visualizations of different types using `plot_ly()` and `ggplotly()`.
- Customize the hoverinfo and other plot features.
- Create a Choropleth map using `plot_geo()`.

# Lab Description

We will work with two Starbucks datasets, one on the store locations (global) and one for the nutritional data for their food and drink items. We will do some text analysis of the menu items.

# Deliverables

Upload an html file to Quercus and make sure the figures remain interactive.

# Steps

### 0. Install and load libraries

```{r message=FALSE, echo=FALSE, warning=FALSE}

library(tidyverse)
library(plotly)
library(widgetframe)
library(tidytext)

```

### 1. Read in the data

- There are 4 datasets to read in, Starbucks locations, Starbucks nutrition, US population by state, and US state abbreviations. All of them are on the course GitHub.

```{r}

sb_locs <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/refs/heads/main/data/starbucks/starbucks-locations.csv")

sb_nutr <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/refs/heads/main/data/starbucks/starbucks-menu-nutrition.csv")

usa_pop <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/refs/heads/main/data/starbucks/us_state_pop.csv")

usa_states<-read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/refs/heads/main/data/starbucks/states.csv")
```

### 2. Look at the data

- Inspect each dataset to look at variable names and ensure it was imported correctly.


```{r}

```

### 3. Format and merge the data

- Subset Starbucks data to the US.
- Create counts of Starbucks stores by state.
- Merge population in with the store count by state.
- Inspect the range values for each variable. 

```{r}

sb_usa <- sb_locs |> filter()

sb_locs_state <- sb_usa |>
  group_by() |>
  rename() |>
  summarize()

# need state abbreviations
usa_pop_abbr <- 
  full_join(
            ) 
  
sb_locs_state <- full_join()
  


```

### 4. Use `ggplotly` for EDA

Answer the following questions:

- Are the number of Starbucks proportional to the population of a state? (scatterplot)

- Is the caloric distribution of Starbucks menu items different for drinks and food? (histogram)

- What are the top 20 words in Starbucks menu items? (bar plot)

```{r}



```

- 4a) Answer: 

```{r}



```

- 4b) Answer: 

```{r}


```

- 4c) Answer: 

### 5. Scatterplots using `plot_ly()`

- Create a scatterplot using `plot_ly()` representing the relationship between calories and carbs. Color the points by category (food or beverage). Is there a relationship, and do food or beverages tend to have more calories?

```{r}

```


- 5a) Answer:

- Repeat this scatterplot but for the items that include the top 10 words. Color again by category, and add hoverinfo specifying the word in the item name. Add layout information to title the chart and the axes, and enable `hovermode = "compare"`.
- What are the top 10 words and is the plot much different than above?
  
```{r}

  
```

- 5b) Answer:



### 6. `plot_ly` Boxplots

- Create a boxplot of all of the nutritional variables in groups by the 10 item words.
- Which top word has the most calories? Which top word has the most protein?

```{r}


```

- 6) Answer:

### 7. 3D Scatterplot

- Create a 3D scatterplot between Calories, Carbs, and Protein for the items containing the top 10 words
- Do you see any patterns (clusters or trends)?

```{r}


```

- 7) Answer:


### 8. `plot_ly` Map

- Create a map to visualize the number of stores per state, and another for the population by state. Add custom hover text. Use subplot to put the maps side by side.
- Describe the differences if any.

```{r}
# Set up mapping details
set_map_details <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('steelblue')
)

# Make sure both maps are on the same color scale
shadeLimit <- 125

# Create hover text
sb_locs_state$hover <- with(sb_locs_state, paste("Number of Starbucks: ", n_stores, '<br>', "State: ", state.y, '<br>', "Population: ", population))

# Create the map
map1 <- plot_geo()
map1


map2 <- plot_geo()
map2

subplot(map1, map2)
```

- 8) Answer:
