---
title: "Lab 03 - Exploratory Data Analysis"
date: "January 22, 2025"
---

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
#install.packages(c("data.table","tidyverse",leaflet", "reticulate", "lubridate"))
library(data.table)
library(leaflet)
library(tidyverse)
library(lubridate)
#library(reticulate) #if you want to use Python code chunks
```

# Learning Goals

- Read in and get familiar with the meteorology dataset  
- Step through the EDA "checklist" presented in the class slides
- Practice making exploratory graphs

# Lab Description

We will work with the meteorological data presented in lecture. Recall the dataset consists of weather station readings in the contiguous US. 

**The objective of the lab is to find the weather station with the highest elevation and look at patterns in the time series of its wind speed and temperature.**

# Steps

### 1. Read in the data

First download and then read in with `data.table` function `fread()`


```{r, echo=TRUE, message=FALSE, eval=FALSE}
download.file(
  "https://raw.githubusercontent.com/JSC370/JSC370-2025/main/data/met/met_all.gz",
  destfile = "met_all.gz",
  method   = "curl",
  timeout  = 60
  )

met <- data.table::fread("met_all.gz")
```

### 2. Check the dimensions, headers, footers. How many columns, rows are there?

- Note you may use R or Python

```{r}
# <place your code here>
```

- _Summarize here_

### 3. Take a look at the variables in the dataset. What kind of variables are there?

```{r}
# <place your code here>
```

- _Summarize here_


### 4. Take a closer look at the key variables. 

- Are there missing data? If so, make sure they are coded correctly. 
- Are there any unusual values that look suspicious? Recall the temperature values in the lecture.


```{r}
# <place your code here>
```

- _Summarize here_


### 5. Check the data against an external data source and make adjustments.

- Check that the range of elevations make sense.
- Google or ChatGPT is your friend here.
- Fix any problems that arise in your checks.

```{r}
# <place your code here>
```

- _Summarize anything that was removed_


### 6. Calculate summary statistics

Remember to keep the initial question in mind. We want to pick out the weather station with maximum elevation and examine its wind speed and temperature.

Some ideas: 
1. select the weather station with maximum elevation; 
2. look at the correlation between temperature and wind speed; and 
3. look at the correlation between temperature and wind speed with hour and day of the month.

```{r}
# <place your code here>
```

### 7. At what elevation is the highest weather station?

```{r}
# <place your code here>
```

- _Summarize here_

### 8. Exploratory graphs: Distributions

We should look at the distributions of all of the key variables (elevation, temp, wind speed) to make sure there are no remaining issues with the data.

```{r }
# <place your code here>
```

One thing we should consider for later analyses is to log transform wind speed and elevation as the are very skewed.

- _Summarize_

### 9. Exploratory graphs: Time Series

Look at the time series of temperature and wind speed at this location. For this we will need to create a date-time variable for the x-axis.

```{r}
# <place your code here>
```

With the date-time variable we can plot the time series of temperature and wind speed. Summarize any trends that you see in these time series plots.

```{r}
# <place your code here>
```

- _Summarize_

### 10. Exploratory graphs: Map

Where is the weather station with highest elevation located? (i.e. make a map!)

```{r}
# <place your code here>
# hint: make use of leaflet
```