---
title: "Lab 9 - HPC"
output: 
html_document: default
link-citations: yes
---

# Learning goals

In this lab, you are expected to practice the following skills:

- Evaluate whether a problem can be parallelized or not.
- Practice with the parallel package.
- Use Rscript to submit jobs.

```{r eval=FALSE, echo=FALSE}
# install any missing packages
install.packages("microbenchmark")
```

## Problem 1

Give yourself a few minutes to think about what you learned about parallelization. List three
examples of problems that you believe may be solved using parallel computing,
and check for packages on the HPC CRAN task view that may be related to it.

Monte Carlo Simulations - Many iterations of random sampling can be parallelized. Related package: parallel, foreach.

Genome-Wide Association Studies (GWAS) - Large-scale computations on genetic data can be parallelized. Related package: BiocParallel.

Image Processing (e.g., edge detection, filtering) - Parallel processing of image pixels can speed up performance. Related package: future.apply.

## Problem 2: Pre-parallelization

The following functions can be written to be more efficient without using
`parallel`:

1. This function generates a `n x k` dataset with all its entries having a Poisson distribution with mean `lambda`.

```{r p2-fun1, eval = TRUE}
fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  
  for (i in 1:n)
    x <- rbind(x, rpois(k, lambda))
  
  return(x)
}

fun1alt <- function(n = 100, k = 4, lambda = 4) {
 matrix(rpois(n * k, lambda), nrow = n, ncol = k)
}

# Benchmarking
microbenchmark::microbenchmark(
  fun1(),
  fun1alt()
)
```

How much faster?

Around 127 microseconds faster (on average)


2.  Find the column max (hint: Checkout the function `max.col()`).

```{r p2-fun2, eval = TRUE}
# Data Generating Process (10 x 10,000 matrix)
set.seed(1234)
x <- matrix(rnorm(1e4), nrow=10)

# Find each column's max value
fun2 <- function(x) {
  apply(x, 2, max)
}

fun2alt <- function(x) {
  x[max.col(t(x), ties.method = "first") + (0:(ncol(x)-1))*nrow(x)]
}

# Benchmarking
library(ggplot2)
library(microbenchmark)

benchmark_results <- microbenchmark(
  fun2(x),
  fun2alt(x)
)

# Plotting the benchmark results
plot(benchmark_results)
```

_Answer here with a plot._


## Problem 3: Parallelize everything

We will now turn our attention to non-parametric 
[bootstrapping](https://en.wikipedia.org/wiki/Bootstrapping_(statistics)).
Among its many uses, non-parametric bootstrapping allow us to obtain confidence
intervals for parameter estimates without relying on parametric assumptions.

The main assumption is that we can approximate many experiments by resampling
observations from our original dataset, which reflects the population. 

This function implements the non-parametric bootstrap:

```{r p3-boot-fun, eval = TRUE}
  library(parallel)

my_boot <- function(dat, stat, R, ncpus = 1L) {
  
  # Getting the random indices
  n <- nrow(dat)
  idx <- matrix(sample.int(n, n*R, TRUE), nrow=n, ncol=R)

  # Making the cluster using `ncpus`
  cl <- makeCluster(ncpus)
  clusterExport(cl, varlist = c("dat", "stat", "idx"), envir = environment())
  
  # Running parallel bootstrapping
  ans <- parLapply(cl, seq_len(R), function(i) {
    stat(dat[idx[,i], , drop=FALSE])
  })
  
  stopCluster(cl)
  
  # Coercing the list into a matrix
  ans <- do.call(rbind, ans)
  
  ans
}
```

1. Use the previous pseudocode, and make it work with `parallel`. Here is just an example for you to try:

```{r p3-test-boot, eval = TRUE}
# Bootstrap of a linear regression model
my_stat <- function(dat) {
  coef(lm(y ~ x, data = dat))
}

# DATA SIM
set.seed(1)
n <- 500
R <- 1000
x <- rnorm(n)
y <- x*5 + rnorm(n)
dat <- data.frame(x, y)

# Check if we get something similar as lm
ans0 <- confint(lm(y ~ x, data = dat))
ans1 <- my_boot(dat, my_stat, R, ncpus = 2)

# Checking bootstrap results
apply(ans1, 2, quantile, probs = c(0.025, 0.975))
```

2. Check whether your version actually goes faster than the non-parallel version:

```{r benchmark-problem3, eval = TRUE}
microbenchmark::microbenchmark(
  my_boot(dat, my_stat, R, ncpus = 1),
  my_boot(dat, my_stat, R, ncpus = 2)
)
```

It takes about 30 milliseconds less, on average.

## Problem 4: Compile this markdown document using Rscript

Once you have saved this Rmd file, try running the following command
in your terminal:

```bash
Rscript --vanilla -e 'rmarkdown::render("[full-path-to-your-Rmd-file.Rmd]")' &
```

Where `[full-path-to-your-Rmd-file.Rmd]` should be replace with the full path to
your Rmd file... :).


